---
title: "project"
author: "Sirapu Nandini"
date: "2024-11-04"
output:
  pdf_document: default
  html_document: default
---

```{r}
options(repos = c(CRAN = "https://cran.r-project.org"))

```


```{r}
#Data loading 
url <- "https://raw.githubusercontent.com/srirapunandini/dav---5400/refs/heads/main/StudentPerformanceFactors%20(2).csv"

# Reading the CSV file into R
project_data <- read.csv(url)

```
Consists of 6,607 rows and 20 columns.


```{r}
# View summary statistics of the dataset
summary(project_data)
```

```{r}
#plots for the variables (numerical and categorical)
library(ggplot2)

# Loop through each variable and create plots
for (var in names(project_data)) {
  if (is.numeric(project_data[[var]])) {
    # Histogram for numeric variables
    p <- ggplot(project_data, aes(x = .data[[var]])) +
      geom_histogram(binwidth = 1, fill = "blue", color = "black") +
      ggtitle(paste("Histogram of", var)) +
      theme_minimal() +
      xlab(var) +
      ylab("Frequency")
  } else {
    # Bar plot for categorical variables
    p <- ggplot(project_data, aes(x = .data[[var]])) +
      geom_bar(fill = "skyblue", color = "black") +
      ggtitle(paste("Bar Plot of", var)) +
      theme_minimal() +
      xlab(var) +
      ylab("Count")
  }
  print(p)
}


```
Histograms for Numeric Variables: Hours Studied, Attendance, Sleep Hours, Previous Scores, Exam Scores. The histograms for these variables show a right-skewed distribution, meaning most students fall into lower-to-mid ranges for hours studied, attendance, and exam scores. We found that this is typical for educational datasets, where most students show average to below-average performance, with fewer students at the extreme high end. This suggests that while many students study for a moderate number of hours, there are a few who might be over-preparing or under-preparing for exams.
Bar Plots for Categorical Variables: Parental Involvement, Access to Resources, Motivation Level, Family Income, School Type, Teacher Quality: These variables are mostly categorical, showing high counts in medium and high categories for involvement, resources, and motivation. Income is skewed towards medium, while teacher quality has a balanced distribution. Most students have high parental involvement, access to resources, and motivation. The high counts in the "medium" categories for family income and teacher quality suggest that these factors are central to the student population
 
```{r}
# Check for null values in each column
#Data cleaning checking for null values 
colSums(is.na(project_data))

```
we found that there are zero null values.
```{r}
#using Z_scores for checking the outliers 
z_scores <- scale(project_data[sapply(project_data, is.numeric)])
outliers <- abs(z_scores) > 3
colSums(outliers)

```
There are outliers in the hours_studied , Tutoring_sessions and the Exam_score columns.

```{r}
#plots showing the outliers
library(ggplot2)

# Variables with outliers
outlier_vars <- c("Hours_Studied", "Attendance", "Sleep_Hours", "Previous_Scores", 
                  "Tutoring_Sessions", "Physical_Activity", "Exam_Score")

# Create boxplots for variables with outliers
for (var in outlier_vars) {
  p <- ggplot(project_data, aes(y = .data[[var]])) +
    geom_boxplot(fill = "lightblue") +
    ggtitle(paste("Boxplot of", var)) +
    theme_minimal() +
    ylab(var)
  print(p)
}



```
Outliers:Boxplots for Hours Studied, Attendance, Tutoring Sessions, Sleep Hours: Some outliers were identified, especially in hours studied, tutoring sessions, and exam scores.A few students are studying far more (or less) than their peers, and similarly, a small number of students might have abnormal exam scores compared to the rest. These outliers could represent exceptional or struggling students.


```{r}
#using Z_scores to remove the outliers 
z_scores <- scale(project_data[sapply(project_data, is.numeric)])
project_data <- project_data[apply(z_scores, 1, function(x) all(abs(x) <= 3)), ]

```
 
```{r}
#plots to remove the outliers 
library(ggplot2)

# Create boxplots again for variables after outliers removal
for (var in outlier_vars) {
  p <- ggplot(project_data, aes(y = .data[[var]])) +
    geom_boxplot(fill = "lightblue") +
    ggtitle(paste("Boxplot of", var, "after removing outliers")) +
    theme_minimal() +
    ylab(var)
  print(p)
}

```
Still we can find the outliers in the particular columns like hours_studied , Tutoring_sessions and the Exam_score.
```{r}
#using IQR (INTER QUARTILE RANGE) to remove the outliers
for (var in outlier_vars) {
  Q1 <- quantile(project_data[[var]], 0.25)
  Q3 <- quantile(project_data[[var]], 0.75)
  IQR <- Q3 - Q1
  project_data <- project_data[!(project_data[[var]] < (Q1 - 1.5 * IQR) | project_data[[var]] > (Q3 + 1.5 * IQR)), ]
}

```
 
```{r}
#plots to remove outliers
library(ggplot2)

# Create boxplots again for variables after outliers removal
for (var in outlier_vars) {
  p <- ggplot(project_data, aes(y = .data[[var]])) +
    geom_boxplot(fill = "lightblue") +
    ggtitle(paste("Boxplot of", var, "after removing outliers")) +
    theme_minimal() +
    ylab(var)
  print(p)
}

```
1. Histogram of Hours Studied:
The histogram shows that most students studied between 10 and 30 hours, with a few students dedicating significantly more or fewer hours to studying, indicating a skew towards moderate study times.
2. Histogram of Attendance:
The attendance histogram reveals that the majority of students have high attendance rates, with most attending between 80% and 100% of the classes, reflecting a generally engaged student body.
3. Histogram of Sleep Hours:
The distribution of sleep hours shows that most students sleep between 6 and 8 hours per night, with few extreme values at either end, indicating a typical range of sleep habits.
4. Histogram of Exam Scores:
The exam score histogram displays a relatively normal distribution, with most students scoring between 60 and 80, suggesting a concentration around the average exam performance.
5. Bar Plot of Parental Involvement:
The bar plot demonstrates that most students have medium to high parental involvement, with fewer students reporting low involvement, highlighting the importance of family support in student performance.
6. Bar Plot of Extracurricular Activities:
The bar plot shows that the majority of students engage in extracurricular activities, with a smaller proportion reporting no participation, indicating a positive correlation between involvement outside of class and academic life.
7. Boxplot of Exam Scores by Motivation Level:
The boxplot illustrates that students with higher motivation levels tend to score higher on exams, with the lower motivation group showing a wider spread in exam scores.
8. Boxplot of Exam Scores by Internet Access:
The boxplot highlights that students with access to the internet generally have higher exam scores, with more consistent results across this group compared to those without internet access.

```{r}
install.packages("corrplot")
library(corrplot)

```
```{r}
install.packages("corrplot", repos = "https://cran.r-project.org")

```



```{r}
# Assuming 'project_data' is your data frame
# Load necessary libraries
library(ggplot2)
library(dplyr)

# 1. Univariate Analysis
# Histogram of Exam_Score
ggplot(project_data, aes(x = Exam_Score)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Exam Scores", x = "Exam Score", y = "Frequency")

# Summary statistics for Hours_Studied
summary(project_data$Hours_Studied)

# Bar plot for Parental Education Level
ggplot(project_data, aes(x = Parental_Education_Level)) +
  geom_bar(fill = "salmon", color = "black") +
  labs(title = "Parental Education Level Distribution", x = "Parental Education Level", y = "Frequency")

# 2. Bivariate Analysis
# Scatter plot of Exam_Score vs. Hours_Studied
ggplot(project_data, aes(x = Hours_Studied, y = Exam_Score)) +
  geom_point(color = "purple") +
  labs(title = "Exam Score vs. Hours Studied", x = "Hours Studied", y = "Exam Score")

# Box plot of Exam_Score by Gender
ggplot(project_data, aes(x = Gender, y = Exam_Score)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Exam Score by Gender", x = "Gender", y = "Exam Score")

# Correlation matrix for numerical features
# Selecting numeric columns for correlation
num_data <- project_data %>% select(Exam_Score, Hours_Studied, Attendance, Previous_Scores)
cor_matrix <- cor(num_data, use = "complete.obs")

# Display correlation matrix
library(corrplot)
corrplot::corrplot(cor_matrix, method = "color", addCoef.col = "black", tl.col = "black", title = "Correlation Matrix", mar = c(0, 0, 1, 0))

```
.Correlation Heatmap of Numeric Variables:
The correlation heatmap shows a strong positive relationship between hours studied, attendance, and exam scores, suggesting that both study habits and class attendance are key predictors of academic success..The picture shows a correlation matrix, which tells us how different factors relate to exam scores. "Attendance" and "Hours Studied" have a strong positive relationship with "Exam Score" (0.68 and 0.50, respectively), meaning students who attend more and study more tend to score higher. "Previous Scores" has a weaker relationship with "Exam Score" (0.20).
. Scatter Plot of Hours Studied vs Exam Scores:
The scatter plot demonstrates a positive linear relationship between hours studied and exam scores, with students who study more generally achieving higher scores.

.This histogram shows the distribution of exam scores in the dataset. The majority of scores fall between 65 and 70, with fewer students scoring below 60 or above 75. This indicates that most students' performance is clustered around the average, with fewer outliers.

.The boxplot shows exam scores by gender. Both female and male students have similar exam score ranges, with medians around 68-70. There doesn’t appear to be a big difference in exam scores between genders.
.This scatter plot shows the relationship between hours studied and exam scores. Generally, as students study more hours, their exam scores tend to increase. The dots form an upward pattern, showing a positive link between studying time and exam performance.

 
```{r}
# Fit the linear regression model
model <- lm(Exam_Score ~ ., data = project_data)

# Summarize the model to see the significance of each factor
summary(model)

```
 Coefficients:The estimate values shows how each variable impacts exam_score. For example: Hours_studied has a positive coefficient(0.298) indicating that more study hours are associated with a higher score.
 Significance Levels: Most predictors are highly significant (indicated by *** in the Pr(>|t|) column), suggesting they meaningfully impact Exam_Score.
Model Fit: The high R-squared (0.99) indicates that this model explains about 99% of the variance in Exam_Score, suggesting a strong fit.
 